Probability Aggregation: Dynamic Hierarchical Modeling of Expert Beliefs in Geopolitical Events

### 2013-08-24 paper.tex
Edits

\section{Introduction}
\label{intro}
Individual experts can differ radically from one another in their abilities to assess probabilities of future events. The quality of a probability assessment is typically evaluated in regards to an important attribute called \textit{calibration}. An expert is said to be well-calibrated if his probabilities match with the long-term frequencies of the events. This means that, for instance, the proportion of events that actually occurred is 60\% for all those events for which he assessed a probability of 0.60. Even though several experiments have been conducted to show that experts are generally poorly calibrated (see, e.g., \citet{cooke1991experts, shlyakhter1994quantifying}), expertise has been found to be associated with calibration. In particular, \citet{wright1994coherence} argue that a higher level of self-reported expertise on the subject matter implies better calibration.

Unfortunately, calibration by itself does not guarantee usefulness. To see this, consider a relatively stationary process, such as rain in some geographic region, whose empirical frequency of occurrence in the last 10 years is 45\%. In this setting an expert can assess a constant probability of 0.45 and be well-calibrated. This probability, however, is close to 0.5 and hence too uninformative to be very useful in decision making. For the sake of comparison, consider another expert whose typical probability assessment is near 0.0 or 1.0. In the forecasting literature such an expert is called \textit{sharp}. If the expert is not only sharp but also well-calibrated, then he is able to forecast the behavior of the process both with high certainty and accuracy. From the decision-maker's perspective, his advice is highly useful as it involves very little risk and much information. For these reasons, a common goal in probability estimation is to maximize sharpness subject to calibration (see, e.g., \citet{raftery2005using, Ranjan08}). This is a well-defined goal that has led to a wide range of novel and insightful observations in probability forecasting literature.

@@

%If the consulted experts diligently make probability assessments on a daily basis, many previously-studied time-series procedures could be used to aggregate their assessments.

This paper adds another layer of complexity by assuming a heterogeneous set of experts, most of whom only make one or two probability assessments over the hundred or so days before the event resolves. The problem at hand is therefore strikingly different from many time-series estimation problems, where one has an observation at every time point -- or almost every time point.
Therefore standard time-series procedures like ARIMA (see, e.g., \citet{mills1991time}) are not applicable without ignoring the heterogeneity among the experts. In response to the lack of procedures to model sparse and heterogeneous probability forecasting data, this paper introduces an interpretable time-series model that incorporates self-reported expertise and captures a sharp and well-calibrated estimate of the crowd belief. The model is then used for
\begin{itemize}
\item the analysis of group-level under- and overconfidence across different levels of self-reported expertise,
%\item confidence bands for the crowd belief,
\item accurate probability forecasts, and
\item many question-specific quantities that have easy interpretations and can be used to gain novel insight in the social sciences. 
\end{itemize}

@@

\section{Discussion}
This paper began with an introduction of a rather unorthodox but nonetheless realistic probability forecasting setting, in which forecasts are made very infrequently by a heterogeneous group of experts over a period of time. The resulting data is too sparse to be modeled efficiently with standard time-series methods.  In response to this lack of appropriate methodology, our work 
%expands the literature towards less-studied areas of  probability aggregation by
introduces an interpretable time-series model that incorporates self-reported expertise and captures a sharp and well-calibrated crowd belief across time. The model is estimated in two-steps: The first step samples constrained versions of the model parameters via a Gibbs sampler that only makes use of standard distributions. The second step involves a one-dimensional optimization procedure that transforms the constrained parameter values into their unconstrained counterparts.

@@

\subsection{Summary of Findings}
The model was applied to an unusually large dataset on probability forecasting done by human experts. The estimated crowd belief was found to be sharp and well-calibrated under both in- and out-of-sample probability estimation. This has direct implications  on predictive power and model interpretability. Firstly, the model was shown to outperform other probability aggregators in terms of forecasting ability. Secondly, the crowd belief was used as the no-bias reference point to study the distributions of the bias terms for different expertise-groups. All the groups were found to be under-confident. The under-confidence, however, decreased as the level of self-reported expertise increased. This result is about groups of experts and hence does not conflict with the well-known result of the individuals being over-confident (see, e.g.,  \citet{lichtenstein1977calibration, morgan1992uncertainty, bier2004implications}). Besides making predictions or studying group-level bias, the model can be used to generate estimates of many problem-specific quantities. These quantities have clear interpretations and can be combined in many interesting ways to answer a range of hypotheses about the experts and the questions.

\subsection{Future Directions and Limitations}
 The model can be extended in different ways to meet a wide range of research objectives. For instance, the degree to which the group-level under-confidence decreases towards the resolution of the target event, could be studied by allowing the bias vector to depend on time. The model is also by no means limited to the study of group-level bias at different levels of self-reported expertise. The experts could equally well be partitioned based on other features, such as gender, education, or specialty, to produce a range of novel insight on the level of confidence among different groups of expert forecasters. Similarly, the model could be applied to questions grouped by types, such as economic, domestic, or international, to learn about expert forecasting under a wide variety of problem attributes.

>>>>

\section{Introduction}
\label{intro}
Individual experts can differ radically from one another in their abilities to assess probabilities of future events. The quality of a probability assessment is typically evaluated in regards to \textit{calibration} that measures the degree to which the frequency of occurrence of the events agree with the given probabilities. This means that, for instance, the proportion of events that actually occurred is 60\% for all those events for which a well-calibrated expert assessed a probability of 0.60. Unfortunately, several experiments have been conducted to show that experts are generally poorly calibrated (see, e.g., \citet{cooke1991experts, shlyakhter1994quantifying}). \citet{wright1994coherence}, however, argue that a higher level of self-reported expertise on the subject-matter can be associated with better calibration.

Calibration by itself, however, is not sufficient for a useful probability assessment. To see this, consider a relatively stationary process, such rain on different days in some geographic region, whose empirical frequency of occurrence in the last 10 years is 45\%. In this setting an expert can assess a constant probability of 0.45 and be well-calibrated. This assessment, however, can be made without any subject-matter expertise or special training. For this reason the long-term frequency is often considered as the baseline probability -- a naive assessment that provides the decision-maker very little extra information. A consulting expert should therefore aim to make probability assessments that are as far from the baseline as possible. The extent to which his probabilities differ from the baseline is measured with an important attribute called \textit{sharpness} (\citet{gneiting2008rejoinder, winkler2008comments}). If the expert is both sharp and well-calibrated, he is able to forecast the behavior of the process with high certainty and accuracy. From the decision-maker's perspective, his advice is highly useful as it involves much information and very little risk. Therefore useful probability estimation should aim to maximize sharpness subject to calibration (see, e.g., \citet{raftery2005using, murphy1987general}). This is a well-defined goal that has led to a wide range of novel and insightful observations in probability forecasting.

@@

This paper adds another layer of complexity by assuming a heterogeneous set of experts, most of whom only make one or two probability assessments over the hundred or so days before the event resolves. This means that the decision-maker faces a different group of experts  every day, with only a few experts returning later on for a second round of advice. The problem at hand is therefore strikingly different from many time-series estimation problems, where one has an observation at every time point -- or almost every time point. As a result, standard time-series procedures like ARIMA (see, e.g., \citet{mills1991time}) are not directly applicable to our setting. In response to this lack of appropriate procedures, we introduce an interpretable time-series model that incorporates self-reported expertise and captures a sharp and well-calibrated estimate of the crowd belief. The model is then used for
\begin{itemize}
\item the analysis of group-level under- and overconfidence across different levels of self-reported expertise,
%\item confidence bands for the crowd belief,
\item accurate probability forecasts, and
\item many question-specific quantities that have easy interpretations and can be used to gain novel insight in the social sciences. 
\end{itemize}

@@

\section{Discussion}
This paper began with an introduction of a rather unorthodox but nonetheless realistic time-series setting in which probability forecasts are made very infrequently by a heterogeneous group of experts. The resulting data is too sparse to be modeled efficiently with standard time-series methods.  In response to this lack of appropriate methodology, our work 
introduces an interpretable time-series model that incorporates self-reported expertise and captures a sharp and well-calibrated crowd belief across time. The model is estimated in two-steps: The first step, known as the \textit{sampling step}, samples constrained versions of the model parameters via a Gibbs sampler that only makes use of standard distributions. The second step, known as the \textit{calibration step}, involves a one-dimensional optimization procedure that transforms the constrained parameter values into their unconstrained counterparts.

@@

\subsection{Summary of Findings}
The model was applied to an unusually large dataset on probability forecasting done by human experts. The estimated crowd belief was found to be sharp and well-calibrated under both in- and out-of-sample probability estimation. This has direct implications  on predictive power and model interpretability. Firstly, the model was shown to outperform other probability aggregators in terms of forecasting ability. Secondly, the crowd belief was used as the no-bias reference point to study the distributions of the bias terms for different expertise-groups. All the groups were found to be under-confident. The under-confidence, however, decreased as the level of self-reported expertise increased. This result is about groups of experts and hence does not conflict with the well-known result of the individuals being over-confident (see, e.g.,  \citet{lichtenstein1977calibration, morgan1992uncertainty, bier2004implications}). Besides making predictions or studying group-level bias, the model can be used to generate estimates of many problem-specific quantities. These quantities have clear interpretations and can be combined in many interesting ways to answer a range of hypotheses about different types of questions and expert behavior.

\subsection{Future Directions and Limitations}
This paper developed a time-series model that is parsimonious subject to addressing the main challenges that arise from sparse probability forecasting data. The model can be extended in various ways using standard techniques. For instance, most of the model parameters were assumed constant over time. It is intuitively reasonable, however, to assume that the behavior of the model parameters is different a hundred days before as compared to a day before the event resolves.  The evolution of the parameters over time could be studied by allowing the parameters to depend on the time index.

Similarly, the model parameters can be allowed to depend on a variety of expert characteristic. In this paper the multiplicative bias was assumed to depend on the level of self-reported expertise. Similar dependencies, however, could be constructed based on any expert characteristic, such as sex, education, specialty, or even a combination of characteristics, to produce a range of novel insight on the level of confidence across different groups of experts.
%The model is also by no means limited to the study of group-level bias at different levels of self-reported expertise. The experts could equally well be partitioned based on other features, such as gender, education, or specialty, to produce a range of novel insight on the level of confidence among different groups of expert forecasters.
Furthermore, instead of grouping the experts, the questions could be grouped by types, such as economic, domestic, or international. These groups could be analyzed separately to learn about forecasting behavior under a wide variety of problem attributes.

%The model could be applied to questions grouped by types, such as economic, domestic, or international, to learn about expert forecasting under a wide variety of problem attributes.
